waRne - Putting cricket analytics in a spin

### 2020-10-02 paper_rob_working-.Rmd
push
push

---
title: "waRne - Putting cricket analytics in a spin"
bibliography: references.bib
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
authors: Robert(1), James (2)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(out.width = "75%",
                      dev = "png",
                      fig.align = "center",
                      echo = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(stringr)
library(curl)
library(tidyverse)
library(magrittr)
library(lubridate)
library(ggplot2)
library(ggalt)
library(gganimate)
library(gifski)
player_database <- 
  read_csv("/Users/James/James/Cricket_Thesis/Data/new_player_database_crickjam.csv") %>% 
  dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
  distinct() %>% 
  dplyr::filter(player_name != "Rohit Gurunath Sharma") %>% 
  bind_rows(read_csv("/Users/James/James/Cricket_Thesis/Data/new_player_database_crickjam.csv") %>% 
              dplyr::filter(player_id == 34102) %>% 
              dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
              drop_na() %>% 
              distinct()) %>% 
  dplyr::filter(player_name !="Sultan Ahmed")

bowler_vars <- 
  player_database %>% 
  dplyr::select(-c(batting_style, player_name)) %>% 
  drop_na()%>% 
  mutate(pace_spin = bowling_style, "Right-arm fast" = "Pace",
                                "Slow left-arm orthodox"= "Spin",
                                "Legbreak"= "Spin",
                                "Right-arm medium"= "Pace",
                                "Left-arm medium-fast"= "Pace",
                                "Right-arm offbreak"= "Spin",
                                "Legbreak googly"= "Spin",
                                "Left-arm fast"= "Pace",
                                "Right-arm medium-fast"= "Pace",
                                "Left-arm fast-medium"= "Pace",
                                "Right-arm fast-medium"= "Pace",
                                "Left-arm medium"= "Pace",
                                "Slow left-arm chinaman"= "Spin",
                                "Right-arm offbreak, Legbreak"= "Spin",
                                "Right-arm medium, Right-arm offbreak"= "Spin",
                                "Left-arm medium, Slow left-arm orthodox, 
                                               Slow left-arm chinaman"= "Spin", 
                                "Left-arm medium, Slow left-arm orthodox, Slow left-arm chinaman" = "Spin")

@@

cricjam_data <- 
  read_csv("/Users/James/James/Reproducibility-in-Cricket/Data/combined_2020_09_11.csv") %>% 
  dplyr::filter(!grepl("No result", match_summary)) %>%
  mutate(bucketed_year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
  left_join(batting_vars, by = c("batsman_id" = "player_id")) %>% 
  left_join(bowler_vars, by = c("bowler_id" = "player_id")) %>% 
  mutate(batsman_name = str_remove(batsman_name, "\\(\\)"), 
         batsman_name = str_trim(batsman_name), 
         runs = extra_runs+earned_runs)
# 
# cricjam_data_partner <- 
#   cricjam_data %>% 
#   dplyr::select(match_id, inning, batsman_name, batsman_id, batting_order) %>% 
#   distinct() %>% 
#   arrange(batting_order) %>% View()

@@

\color{red}{\text{Yeah I forgot I wouldn't have any internet on the plane, so everything I wrote is not referencing anything from a link or package I didn't have downloaded prior to boarding}}

@@

The importance of reproducibility, and the related issue of open access to data, has received a lot of recent attention. Momentum on these issues is gathering in the sports analytics community, particularly in North America where open access to quality sports data has become the norm. Cricket is the second largest commercial sport in the world, but unlike other popular international sports, there has been no mechanism for the public to access comprehensive statistics on players and teams. Expert commentary currently relies heavily on data that isn't made readily accessible and this produces an unnecessary barrier for the development of an inclusive sports analytics community.

@@

Cricket is the second largest sport in the world. However, unfortunetaly there is no easy accessible way to access ball by ball data nor aggregated statistics of teams and players. Data while available on sites like [espncricinfo](https://www.espncricinfo.com/) are not in an easy to use form. For example, each match is listed on different webpages so hours upon hours of time would be required to copy and paste a single season. Hence, there are significant logistical barriers for prospective fans and analysts studying the game, which stagnats understanding of cricket. The Duckworth-Lewis-Stern table, used to reset targets for matches that experience rain delays in all of cricket, is maintained by one man in Australia. Not even teams have access to the table. \color{red}{text{what do they do, call it into the ICC to get a temporary copy or something when there's a delay?}}

This paper describes the waRne package, the first to provide free and easy access to data for cricket for fans. Web scraping tools are avaiable for fans to easily scrape the play by play commentary data on espncricinfo. For the first time fans can evaluate their favourite teams and players and do so in a reproducible and accessible manner. We have included the data for ODI and T20 from the 2010 archived season on ESPCcricinfo up to the matches on September 8, 2020. If a person wishes to expand this data to include more recent games, this package can scrape and update the dataset.

@@

Data accessibility enables fans, analysts and researchers to better understand the game. Through being able to reproduce common popular metrics, visualisations and article findings.

@@

\color{red}{\text{I'm not sure if we should slam the datajam competition directly. I think the guy tweeting out conflicting news is a good point and I want to keep that in because he was a dick about his results, but we just lost to Quantium and I worry it will make us seem petty or look like sore losers. We could just indirectly show that they were wrong, but not mention them at all, I would feel fine with that}}

For example in this [series of tweets](https://twitter.com/benjonescricket/status/1219898578719723520) we see the narrative being pushed that Steve Smith is a good player vs pace bowling unfortunetly just a few months prior the same company and journalist published an article which had Steve Smith doing much worse against pace (balls above 140km/h). Unlike a similar sport baseball, fans have no easily accisible way of seeing if Steve Smith vs pace is a strength as alluded to in the original tweet, or a weakness like the same persons published online article. In comparsion, fans are able to get a breakdown of Mike Trout vs fastballs from using [baseballr](https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/) which provides access through statcast data from [baseballsavant]([https://baseballsavant.mlb.com)](https://baseballsavant.mlb.com)).

Unfortunetly this is just one of many examples whereby a relatively simple statistic is provided by media and fans have no mechanism to fact check. Fact checking is an important avenue for fans to not only engage and understand statistics, but having this mechanism also stops analytics companies from putting out misleading conclusions and findings.

With the commentary data on cricinfo, we are not able to get the same resolution on delivery speed and location as the statcast data provides in baseball, but we can often learn the type of delivery the batsman faces through the years. This represents an enormous leap in publically available data, and we hope will lead to the eventual release of the ball tracking data used butnot shared in crickets leagues around the world

Armed with this new dataset and the waRne package, we are able to not only compare Steve Smiths performance vs bowling type but we are able to easily compare Steve Smiths performance vs bowling types with other players, across multiple seasons, leagues, and cricket formats.

@@

James here need to figure out the code, do we keep it seperate so in the example we still have to join on bowling type to the ball by ball, or will it just be an extra column joined already.

\color{red}{\text{So the way the scraper is set up, I grab the play links when scraping the scorecard and commentary along with the id and name, and then run a separate scraper for the batsman/bowler profiles. So they are separate datasets that have to be joined together, and we drop mostly everything except batting and bowling style. I can probably make an all in one scraper, but it'll be a bit tricky and if we want to put this out Friday it might be best to leave them separate and just walk people through it in the readme. Otherwise it might not work at all, and the scraper already breaks down so often you have to pay attention more than you should}}

@@

cricjam_data_big %>% 
	dplyr::filter(batsman_id == 267192, 
								!is.na(bowling_style)) %>% 
	group_by(bowling_style, earned_runs) %>% 
	summarise(total = n()) %>% 
	group_by(bowling_style) %>% 
	mutate(totals = sum(total),
				 percent = total/totals, 
				 earned_runs = as.factor(earned_runs)) %>% 
	ggplot(aes(y = bowling_style, x = percent, color = earned_runs))+
	geom_point() +
	labs(color = "Runs", 
			 x = "% Balls Faced", 
			 y = "Bowling Style", 
			 title = "Steve Smith's Distribution of Runs Against Spin and Pace")

Another reason why is that cricket is a weird sport with a lot of old school rules of thumb that are followed/talked about.

For example it has become popular conjecture that its better that teams should bat second (chase down a total set by the oppoition). Through waRne we can answer such question for the T20 game.

For example we can explore do teams win more often when chasing in the big bash over the history of the big bash, or we can plot a line chart to see if there is a movement towards batting last and winning.

cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result, year) %>% 
	summarise(win_count = n()) %>% 
	ungroup() %>% 
	pivot_wider(names_from = batting_team_result, values_from = win_count) %>% 
	rename(superover = 'super-over') %>% 
	replace_na(list(superover=0)) %>% 
	mutate(total_games = lose+win+superover, 
				 win_percentage = win/total_games, 
				 year = as.factor(year)) %>% 
	ggplot(aes(x = year, y = win_percentage, group = 1)) +
	geom_line(color = 'red')

@@

For example using the above we can not only look at the answers graphically, but we can run a statistical test. The example we would use is instead of testing is a coin biased (different from 50/50) we can test to see if the winning \\% is different from $0.5$ for teams chasing.

@@

dat <- 
	cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result) %>% 
	summarise(win_count = n()) %>% 
	dplyr::filter(batting_team_result != 'super-over')

N = sum(dat$win_count)
wins = dat$win_count[2]

z = (wins/N-0.5)/sqrt((0.5^2)/N)

2*(pnorm(z))

# No eveidence to suggest the 2 inning team wins more

@@

James and I entered a reproducibility competition and the winning entry provided a statistic sort of a gateway into the importance of their ''project'' was was specifically in 2019 Australian Big Bash T20 season. The statistic given was in the 2019 big bash season teams that batted first won only $43$% of games. Thankfully we can try to reproduce that statistic now through the use of waRne

Wikipedia tells us that there were the following:

In the 2018-19 Big Bash League there were 59 matches played. Of which we had 3 DLS results.

| Games                             | Result                        |
|-----------------------------------|-------------------------------|
| 21 DEC 2018 - THUNDER VS STARS    | THUNDER WON BY 15 RUNS (D/L)  |
| 2 FEB 2019 - THUNDER VS SIXERS    | SIXERS WON BY 9 WICKETS (D/L) |
| 8 JAN 2019 HEAT VS THUNDER        | HEAT BY 15 RUNS (D/L)         |
| 17th January 2019 Thunder vs Heat | no result                     |

: Games under consideration in 2018-19 Big Bash

```{r, investigation}
df<-read.csv("data/combined_2020_09_11.csv")

blah<-df%>%
	dplyr::select(duckworth_lewis,year,match_id, tournament_name, tournament_type, team, inning, binary_result)%>%
	dplyr::filter(tournament_type%in% c("Twenty20 Internationals", "Twenty20"  ))%>%
	dplyr::filter(tournament_name%in% c("Twenty20 Big Bash","Big Bash League" ))%>%
									dplyr::filter(inning==1)%>%
									dplyr::group_by()%>%
									distinct()
# dataset doesn't contain duckworth lewis results. 
table(blah$year,blah$binary_result)
```

\`\`\`

So here we can see that we in the big bash season 2018-19 we have a current winning percentage batting first of 22/(22+33) =0.4 but this doesn't include the duckworth lewis games or the washed out completely game.

If we were to add in the duckworth lewis games we get 24/(24+34) which is equal to 0.414. Which isn't the result.

We can also clearly see that in the 20192020 big bash season the team batting first wins 39/(39+20) or 0.66 so what could be going on? Did Quantium mean only games played in the actual year 2019?

In that case from the 2018-2019 season we would remove the following games

| Date                 | Game                    | Result                   | Winning Team |
|----------------------|-------------------------|--------------------------|--------------|
| 19 December 2018     | Heat Vs Strickers       | Strickers 5 Wickets      | Chased       |
| 20th December 2018   | Scorchers Vs Renegades  | Renegades by 4 wickets   | Chased       |
| 21 December 2018     | Thunder vs Stars        | Thunder by 15 runs (D/L) | Batted First |
| 22 December 2018     | Sixers vs Scorchers     | Sixers by 17 runs        | Batted First |
| 22 December          | Hurricanes vs Heat      | Hurricanes by 15 runs    | Batted First |
| 23 December 2018     | Strikers vs Renegades   | Renegades by 5 wickets   | Chased       |
| 24 December 2018     | Stars vs Hurricanes     | Hurricanes by 6 wickets  | Chased       |
| 24th December 2018   | Thunder vs sixers       | Thunder by 21 runs       | Batted First |
| 26th December 2018   | Strickers vs scorchers  | Scorchers by 7 wickets   | Chased       |
| 27th December        | Sixers vs Stars         | stars by 5 wickets       | Chased       |
| 28th December        | thunder vs hurricanes   | hurricanes by 7 wickets  | Chased       |
| 29th december        | sixers vs renegades     | sixers by 33 runs        | Batted First |
| 30th decemeber 2018  | scorchers vs hurricanes | hurricanes by 6 wickets  | Chased       |
| 31st decemember 2018 | strickers vs thunder    | strickers by 20 runs     | Batted First |

: Games in 2018 of the 2018-19 Big Bash Season

i.e. we remove 14 games here of which 6 teams won batting first 16/(22+33-14)

Then from here we have to add on games in the 2019-2020 big bash season that were played in 2019

+--------------------+-------------------------+---------------------------------------+--------------------------+
| Date               | Teams                   | Result                                | Winning Team             |
+====================+=========================+=======================================+==========================+
| 17 December        | Thunder vs Heat         | Thunder by 29 runs                    | Batted First             |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 18th December      | Scorchers vs Sixers     | Sixers by 8 wickets                   | Chased                   |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 19th december      | Renegades vs thunder    | thunder by 6 wickets                  | chased                   |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 20th decemeber     | hurricanes vs sixers    | hurricanes by 25 runs                 | batted first             |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 20th decemember    | stars vs heat           | stars by 22 runs                      | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 21st decemember    | strickers vs thunder    | no result                             | no result                |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 21st december      | scorchers vs renegades  | scorchers by 11 runs                  | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 22 deceember       | stars vs hurricanes     | stars by 52 runs                      | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 22 decemember      | heat vs sixers          | heat by 48 runs                       | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 23 dec             | strickers vs scorchers  | strickers by 15 (DLS)                 | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 24th dec           | renegades vs hurricanes | hurricanes by 7 wickets               | chased                   |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 26 decemember      | sixers vs scorchers     | sixers by 48 runs                     | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 27 dec             | strickers vs stars      | strikers by 5 runs                    | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 28th december      | thunder vs sixers       | sixers won super over (batting first) | sixers by superover      |
|                    |                         |                                       |                          |
|                    |                         |                                       | or sixers batting second |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 29th dec           | strikers vs renegades   | strikers by 18 runs                   | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 30th dec           | hurricanes vs stars     | stars by 4 runs (DLS)                 | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+
| 31st december 2019 | thunder vs strikers     | thunder by 3 runs                     | batting first            |
+--------------------+-------------------------+---------------------------------------+--------------------------+

: Games played in 2019 of the 2019-20 Big Bash

12 batting first and 3 batting second

(16+12)/(12+3+22+33-14)

So over the past 2 years (not including DLS) we have

(22+39)/(20+39+33+22)

i.e. \>50% of teams win chasing so what does that mean?

@@

To the surprise of many, its hard to engage cricket fans into the analytics behind the game. Reasons for this are generally centred around reproducibility and explainability to fans.

@@

```{r}
# graph of powerplay

@@

# Dot plot for openers
cricjam_data_big %>% 
  dplyr::filter(inning == 2, 
                batsman_id %in% opening_batsman_of_interest, 
                # batsman_balls_faced_this_match<16, 
                !is.infinite(required_run_rate), 
                over<6) %>% 
  group_by(batsman_name) %>% 
  summarise(change = mean(change_in_requireed_run_rate), 
            total_balls_faced = n()) %>% 
  ggplot(aes(x = total_balls_faced, y = change, label = batsman_name)) +
  geom_point(size = 2) +
  xlim(c(50, 1250))+
  geom_text(vjust = 1.4, color = "black") +
  theme_minimal()+
  labs(x = "Balls Faced", 
       y = "Average Change in Required Run Rate", 
       title = "Top Opening Batsmen T20")
  theme(axis.text = element_text(size = 12), 
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))

@@

```{r}
# middle overs (enter in the middle)

@@

cricjam_data_big %>% 
  dplyr::filter(inning == 2, 
                batsman_id %in% middle_batsman_of_interest, 
                # batsman_balls_faced_this_match<16, 
                !is.infinite(required_run_rate), 
                # min(batsman_balls_faced_this_match)==1,
                over<16, 
                over>=6) %>% 
  group_by(batsman_id, match_id) %>% 
  mutate(min_balls = if_else(min(batsman_balls_faced_this_match)==1, 1, 0)) %>% 
  ungroup() %>% 
  # dplyr::filter(min_balls==1) %>%
  group_by(batsman_name) %>% 
	# dplyr::filter(batsman_id == 28081) %>% ggplot(aes(x = change_in_requireed_run_rate))+geom_histogram()
  summarise(change = mean(change_in_requireed_run_rate), 
            total_balls_faced = n()) %>% 
  ggplot(aes(x = total_balls_faced, y = change, label = batsman_name)) +
  geom_point() +
	xlim(c(300, 1300))+
  # geom_text(vjust = 1, color = "black") +
  labs(x = "Balls Faced", 
       y = "Average Change in Required Run Rate", 
       title = "Top Middle Batsmen T20")+
  geom_text(vjust = 1.4, color = "black") +
  theme_minimal() 
  theme(axis.text = element_text(size = 12), 
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))

@@

@@

@@

@@

pollard %>% 
	bind_rows(maxwell, dhoni, russell) %>%
	# dplyr::filter(bucketed_over <20) %>% 
	ggplot(aes(x = bucketed_over, y = avg, group = batsman_partner, color = batsman_partner)) +
  geom_smooth()+
  geom_hline(yintercept = 0)+
  labs(y = "Required Run Rate", 
       x = "Over", 
       title = "Change in Required Run Rate While Batting (or Partnering)", 
  		 color = "")+
  facet_wrap(~batsman_name)+
  theme_minimal()+
  theme(axis.text = element_text(size = 12), 
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16))

@@

## notes for James

So I don't think its like a complex paper, because of the way cricket is at the moment I think the biggest influence this stuff can make is

-   being able to fact check

-   having batsman metrics that are beyond average, total runs and SR

-   What we can maybe mention? is that change in RRR while simple seems to be quite effective, a better measurement would be the DLS which we know can be derived from the ball by ball data or alternatively if someone has a resource table they can instead of change in RRR use change in resources or whatever

For the examples, I think its pretty easy to get the graphs to look nice, its more about whats the dataset look like? i.e. do we need the first few lines for reproducibility to be joins or is it sort of like done already

>>>>

---
title: "waRne - Putting cricket analytics in a spin"
authors: Robert(1), James (2)
output: 
  pdf_document:
    toc: yes
  html_document:
    toc: true
    toc_float: true
    keep_md: true
bibliography: references.bib
---

```{r setup,message=FALSE, include=FALSE}
knitr::opts_chunk$set(message = TRUE, warning = TRUE)
library(tidyverse)
player_database <- 
  read_csv("data/new_player_database_crickjam (2).csv") %>% 
  dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
  dplyr::distinct() %>% 
  dplyr::filter(player_name != "Rohit Gurunath Sharma") %>% 
  bind_rows(read_csv("data/new_player_database_crickjam (2).csv") %>% 
              dplyr::filter(player_id == 34102) %>% 
              dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
              drop_na() %>% 
              distinct()) %>% 
  dplyr::filter(player_name !="Sultan Ahmed")

bowler_vars <- 
  player_database %>% 
  dplyr::select(-c(batting_style, player_name)) %>% 
  drop_na()%>% 
  mutate(bowling_style = recode(bowling_style, "Right-arm fast" = "Pace",
                                "Slow left-arm orthodox"= "Spin",
                                "Legbreak"= "Spin",
                                "Right-arm medium"= "Pace",
                                "Left-arm medium-fast"= "Pace",
                                "Right-arm offbreak"= "Spin",
                                "Legbreak googly"= "Spin",
                                "Left-arm fast"= "Pace",
                                "Right-arm medium-fast"= "Pace",
                                "Left-arm fast-medium"= "Pace",
                                "Right-arm fast-medium"= "Pace",
                                "Left-arm medium"= "Pace",
                                "Slow left-arm chinaman"= "Spin",
                                "Right-arm offbreak, Legbreak"= "Spin",
                                "Right-arm medium, Right-arm offbreak"= "Spin",
                                "Left-arm medium, Slow left-arm orthodox, 
                                               Slow left-arm chinaman"= "Spin", 
                                "Left-arm medium, Slow left-arm orthodox, Slow left-arm chinaman" = "Spin"))

@@

cricjam_data <- 
  read_csv("data/combined_2020_09_11.csv") %>% 
  dplyr::filter(!grepl("No result", match_summary)) %>%
  mutate(bucketed_year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
  left_join(batting_vars, by = c("batsman_id" = "player_id")) %>% 
  left_join(bowler_vars, by = c("bowler_id" = "player_id")) %>% 
  mutate(batsman_name = str_remove(batsman_name, "\\(\\)"), 
         batsman_name = str_trim(batsman_name), 
         runs = extra_runs+earned_runs)
# 
# cricjam_data_partner <- 
#   cricjam_data %>% 
#   dplyr::select(match_id, inning, batsman_name, batsman_id, batting_order) %>% 
#   distinct() %>% 
#   arrange(batting_order) %>% View()

@@

@@

The importance of reproducibility, and the related issue of open access to data, has received a lot of recent attention.\^[why this not work] Momentum on these issues is gathering in the sports analytics community. While cricket is the second largest commercial sport in the world, unlike other popular international sports, there has been no mechanism for the public to access comprehensive statistics on players and teams. Expert commentary currently relies heavily on data that isn't made readily accessible and this produces an unnecessary barrier for the development of an inclusive sports analytics community.

@@

Cricket is the second largest sport in the world. However, unfortunetaly there is no easy accessible way to access ball by ball data nor aggregated statistics of teams and players. Data while available on sites like [espncricinfo](https://www.espncricinfo.com/) are not in an easy to use form. For example, each match is listed on different webpages so hours upon hours of time would be required to copy and paste a single season, not to mention the added difficulty of linking players between games and competitions in different countries. Hence, there are significant logistical barriers for prospective fans and analysts studying the game, which stagnats understanding of cricket.

This paper describes the waRne package, the first to provide free and easy access to data for cricket for fans. Web scraping tools are avaiable for fans to easily scrape the play by play commentary data on espncricinfo. For the first time fans can evaluate their favourite teams and players and do so in a reproducible and accessible manner. We hope that this package can be used for fans to better understand the game, for teachers to use for interesting examples in class and for analysts in clubland who might not have access to ball by ball data.

@@

Data accessibility enables fans, analyts and researchers to better understand the game. Through being able to reproduce common popular metrics, visualisations and article findings.

@@

For example in this [series of tweets](https://twitter.com/benjonescricket/status/1219898578719723520) we see the narrative being pushed that Steve Smith is a good player vs pace bowling unfortunetly just a few months prior the same company and journalist published an article which had Steve Smith doing much worse against pace (balls above 140km/h). Unlike a similar sport baseball, fans have no easily accisible way of seeing if Steve Smith vs pace is a strength as alluded to in the original tweet, or a weakness like the same persons published online article. In comparsion, fans are able to get a breakdown of Mike Trout vs fastballs from using [baseballr](https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/) which provides access through statcast data from [baseballsavant](https://baseballsavant.mlb.com/).

Unfortunetly this is just one of many examples whereby a relatively simple statistic is provided by media and fans have no mechanism to fact check. Fact checking is an important avenue for fans to not only engage and understand statistics, but having this mechanism also stops analytics people/companies from putting out misleading conclusions and findings.

However though using waRne we are able to not only compare Steve Smiths performance vs bowling type but we are able to easily compare Steve Smiths performance vs bowling types with other players.

@@

Fans of cricket and Australian cricket especially might be interested to see if this is a statistical qwerk of small samples or if Steve Smith generally does perform as an elite cricketer vs pace bowling. Teachers of statistics classes might use this as an interesting example to introduce [tidyverse](https://www.tidyverse.org/learn/) principals [@wickham2019welcome].

```{r}
library(tidyverse)
cricjam_data_big %>% 
	dplyr::filter(batsman_id == 267192, 
								!is.na(bowling_style)) %>% 
	group_by(bowling_style, earned_runs) %>% 
	summarise(total = n()) %>% 
	group_by(bowling_style) %>% 
	mutate(totals = sum(total),
				 percent = total/totals, 
				 earned_runs = as.factor(earned_runs)) %>% 
	ggplot(aes(y = bowling_style, x = percent, color = earned_runs))+
	geom_point() +
	labs(color = "Runs", 
			 x = "% Balls Faced", 
			 y = "Bowling Style", 
			 title = "Steve Smith's Distribution of Runs Against Spin and Pace")

@@

# insert above plots here
```

@@

For example using the above we can not only look at the answers graphically, but we can run a statistical test. A common classroom example for learning the binomial distribution is to ask if a coin is based given a proportion of heads and tails given a sample size [@mcelreath2020statistical]. Instead of asking about coins, fans of sport and cricket might be interested to ask the question; after winning the coin toss, should a team decide to bat first or bat second [@kvam2004teaching].

While a coin toss is a seemingly trivial example, like in most professional sports the winner of the coin toss gets to decide what to do. In cricket, the winner of the coin toss gets to decide if they want to bat first and thus set the total that the opposition needs to pass by a single run to win, or if they want to bowl first and thus chase down the total set. Deciding what to do after winning a coin toss has proved to be a popular media piece in recent years\\footnote{<https://www.espncricinfo.com/story/_/id/21489056/stuart-wark-cricket-move-away-coin-toss> }\\footnote{https[://www.espncricinfo.com/video/clip/\_/id/23230682%5D](https://www.espncricinfo.com/video/clip/_/id/23230682%5D){.uri}}

\\footnote{[https://www.espncricinfo.com/story/\_/id/21489056/stuart-wark-cricket-move-away-coin-toss](https://www.espncricinfo.com/story/_/id/21489056/stuart-wark-cricket-move-away-coin-toss?%5D)}

\^[[https://www.espncricinfo.com/story/\_/id/20429499/lehmann-backs-scrapping-toss]](https://www.espncricinfo.com/story/_/id/20429499/lehmann-backs-scrapping-toss%5D) \^[[https://www.espncricinfo.com/story/\_/id/28770755/how-much-does-losing-tosses-impact-visiting-teams]](https://www.espncricinfo.com/story/_/id/28770755/how-much-does-losing-tosses-impact-visiting-teams%5D) \^[<https://www.forbes.com/sites/tristanlavalette/2018/08/20/are-cricket-matches-being-decided-by-the-luck-of-a-coin-toss/#735456837eff>]

\^[<https://www.statsinsider.com.au/bbl/how-important-is-winning-the-toss-in-the-big-bash-league>]

\^[<https://www.espncricinfo.com/story/_/id/18568387/tim-wigmore-how-batting-second-become-more-fruitful-more-popular>]

Fans and analysts of the game might want to make this decision based on a simple statistic, for example they might make it based on answer these questions instead.

-   Do teams that bat second have a higher winning percentage than those who bat first?

-   Is this consistent across leagues and levels of cricket?

@@

cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result, year) %>% 
	summarise(win_count = n()) %>% 
	ungroup() %>% 
	pivot_wider(names_from = batting_team_result, values_from = win_count) %>% 
	rename(superover = 'super-over') %>% 
	replace_na(list(superover=0)) %>% 
	mutate(total_games = lose+win+superover, 
				 win_percentage = win/total_games, 
				 year = as.factor(year)) %>% 
	ggplot(aes(x = year, y = win_percentage, group = 1)) +
	geom_line(color = 'red')
```

Instead of just looking at this graphically, fans of crickets and teachers of undergraduate statistics might use the dataset as a ''biased coin'' example. So instead of asking, given a proportion of heads over \$n\$ tosses instructors could ask is a team better off batting first or second.

n<-900
p<-0.5
qbinom(c(0.025,0.975),size=n,prob=p)

@@

```{r}
df<-data.frame(
             Year = c(2010L,2011L,2012L,2013L,2014L,
                      2015L,2016L,2017L,2018L,2019L,2020L,2021L,2022L),
  number.of.heads = c(52L,55L,51L,70L,30L,40L,35L,
                      43L,49L,37L,39L,45L,32L)
)

library(broom)
df %>%
  # create variables to pass to binom.test
  mutate(n = 100, 
         x = n - number.of.heads, 
         p = 0.5,
         alternative = "two.sided",
         conf.level = 0.95) %>%
  # Run a binom.test for each row/year
  mutate(result = pmap(list(x, n,p), binom.test)) %>% 
  # Extract results from test into dataframe using broom::tidy
  mutate(result = map(result, tidy)) %>%
  mutate(result = map(result, ~select(.x, p.value, conf.low, conf.high))) %>% 
  unnest(cols = c(result)) %>%
  # plot the annual confidence intervals
  ggplot(aes(factor(Year), p.value)) + 
    geom_errorbar(aes(ymin=conf.low, ymax=conf.high, color = (p.value > 0.05)), width=.1) +
  theme_minimal() +
  labs(y = "P(coin is fair)", x = "Year") +
  ggtitle("Annual coin fairness test")

## James Stein Estimator - Why not a cricket example

<https://bookdown.org/content/922/james-stein.html>

```{r}
```

@@

To the surprise of many, there has not been a real push to engage fans in the analytics of the game of cricket. Unlike other sports which has seen a boom through the use of accessible data like Ice Hockey, NFL, Baseball and Basketball for example.

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE}
# graph of powerplay

@@

# Dot plot for openers
cricjam_data_big %>% 
  dplyr::filter(inning == 2, 
                batsman_id %in% opening_batsman_of_interest, 
                # batsman_balls_faced_this_match<16, 
                !is.infinite(required_run_rate), 
                over<6) %>% 
  group_by(batsman_name) %>% 
  summarise(change = mean(change_in_requireed_run_rate), 
            total_balls_faced = n()) %>% 
  ggplot(aes(x = total_balls_faced, y = change, label = batsman_name)) +
  geom_point(size = 2) +
  xlim(c(50, 1250))+
  geom_text(vjust = 1.4, color = "black") +
  labs(x = "Balls Faced", 
       y = "Average Change in Required Run Rate", 
       title = "Top Opening Batsmen T20")
  theme_minimal()
```

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE}
# middle overs (enter in the middle)

@@

cricjam_data_big %>% 
  dplyr::filter(inning == 2, 
                batsman_id %in% middle_batsman_of_interest, 
                # batsman_balls_faced_this_match<16, 
                !is.infinite(required_run_rate), 
                # min(batsman_balls_faced_this_match)==1,
                over<16, 
                over>=6) %>% 
  group_by(batsman_id, match_id) %>% 
  mutate(min_balls = if_else(min(batsman_balls_faced_this_match)==1, 1, 0)) %>% 
  ungroup() %>% 
  # dplyr::filter(min_balls==1) %>%
  group_by(batsman_name) %>% 
	# dplyr::filter(batsman_id == 28081) %>% ggplot(aes(x = change_in_requireed_run_rate))+geom_histogram()
  summarise(change = mean(change_in_requireed_run_rate), 
            total_balls_faced = n()) %>% 
  ggplot(aes(x = total_balls_faced, y = change, label = batsman_name)) +
  geom_point() +
	xlim(c(300, 1300))+
  # geom_text(vjust = 1, color = "black") +
  labs(x = "Balls Faced", 
       y = "Average Change in Required Run Rate", 
       title = "Top Middle Batsmen T20")+
  geom_text(vjust = 1.4, color = "black") +
  theme_minimal()

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE}

@@

```{r, message=FALSE,echo=FALSE, warning=FALSE}

@@

```{r,message=FALSE,echo=FALSE, warning=FALSE}

@@

pollard %>% 
	bind_rows(maxwell, dhoni, russell) %>%
	# dplyr::filter(bucketed_over <20) %>% 
	ggplot(aes(x = bucketed_over, y = avg, group = batsman_partner, color = batsman_partner)) +
  geom_smooth()+
  geom_hline(yintercept = 0)+
  labs(y = "Required Run Rate", 
       x = "Over", 
       title = "Change in Required Run Rate While Batting (or Partnering)", 
  		 color = "")+
  facet_wrap(~batsman_name)+
  theme_minimal()

@@
